---
title: "CH3"
author: "Nick Oddo"
date: "2024-08-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 3.1 
Sample survey: Suppose we are going to sample 100 individuals from
a county (of size much larger than 100) and ask each sampled person
whether they support policy $Z$ or not. Let $Y_i = 1$ if person $i$ in the sample
supports the policy, and $Y_i = 0$ otherwise.

a. Assume $Y_1,\ldots, Y_{100}$ are, conditional on $\theta$, i.i.d. binary random variables with expectation $\theta$. 
Write down the joint distribution of $Pr(Y_1 = y_1, . . . , Y_{100} = y_{100}|\theta)$ in a compact form.
Also write down the form of $P(\sum Y_i = y|\theta)$.

The joint distribution (likelihood) of $Pr(Y_1 = y_1, . . . , Y_{100} = y_{100}|\theta)$ is:
\begin{align}
 &Pr(Y_1 = y_1, . . . , Y_{100} = y_{100}|\theta) = \prod p^{y_i} \cdot (1-p)^{(1-y_i)}\\
 &=  p^{\sum_{i=1}^{100} y_i} \cdot (1-p)^{(100-\sum_{i=1}^{100} y_i)}

\end{align}


as for $P(\sum Y_i = y|\theta)$, let $Z = \sum Y_i $. Then:
\begin{align}

&P(\sum Y_i = y|\theta) = P(Z = z|\theta)\\
&= \binom{100}{z} p^{z}(1-p)^{100-z}

\end{align}



b. For the moment, suppose you believed that $\theta \in {0.0, 0.1,\ldots , 0.9, 1.0}$.
Given that the results of the survey were $\sum_{i=1}^{100} = 57$ compute $P(\sum Yi = 57|\theta)$ for each of these 11 values of $\theta$ and plot these probabilities as a function of $\theta$.

```{r}

theta = seq(from = 0, to = 1, length.out = 11)
probs = dbinom(x = 57, size = 100,prob = theta)

data.frame(theta,"probs" = round(probs,4))

plot(theta,probs, type = "b")

```


c. Now suppose you originally had no prior information to believe one of
these $\theta$-values over another, and so $Pr(\theta = 0.0) = Pr(\theta = 0.1) = · · · =Pr(\theta = 0.9) = Pr(\theta = 1.0)$.
Use Bayes’ rule to compute $p(\theta|\sum_i=1 Yi =57)$ for each $\theta$-value. Make a plot of this posterior distribution as a
function of $\theta$.

Using a discrete uniform prior $p(\theta) = \frac{1}{11}$, and using $Y = \sum_{i=1}^{100} Y_i$:

\begin{align}

P(\theta \| Y = 57) \\
&= \frac{ P(Y = 57 \| \theta)P(\theta)}{\sum_{j=1}^{11} P(Y = 57\|\theta_j)P(\theta_j) }\\
&=\frac{\binom{100}{57} \theta^57(1-\theta)^43 \cdot \frac{1}{11} }{ \sum_{j=1}^{11} \binom{100}{57}(\frac{1}{11})^57(\frac{10}{11})^43 \cdot (\frac{1}{11})\\

\end{align}

```{r}

# post.probs = dbeta(theta,57+1,100-57+1)/sum(dbeta(theta,57+1,100-57+1))
# 
# post.probs
# 
# data.frame(theta,"Posterior Probs" = round(post.probs,4))
# 
# plot(theta,post.probs,type = 'b')

my_post<-function(theta){ 1/11*(theta/11)^57 * ((1-theta)*10/11)^43}

plot(seq(0,1,length.out = 100),my_post(seq(0,1,length.out = 100)),type = 'l')


# plot(seq(0,1,length.out = 100),pbeta(seq(0,1,length.out = 100),58,44),type = 'l')

```





d. Now suppose you allow θ to be any value in the interval [0, 1]. Using
the uniform prior density for θ, so that p(θ) = 1, plot the posterior
density p(θ) × Pr(Pn
i=1 Yi = 57|θ) as a function of θ.







e. As discussed in this chapter, the posterior distribution of θ is beta(1+
57, 1 + 100−57). Plot the posterior density as a function of θ. Discuss
the relationships among all of the plots you have made for this exercise.

Assuming a flat prior ($\theta \sim Beta(1,1)$) where $f(\theta) = 1 $ the posterior becomes $\theta|\mathcal{D} \sim Beta(y+1,100-y+1)$. So, 

```{r}

thetas.e = seq(0,1,length.out = 900)

post.probs.e = dbeta(thetas.e,57+1,100-57+1)

plot(thetas.e,post.probs.e,type = 'l')

```

