---
title: "A First Course In Bayesian Statistical Methods: CH5"
author: "Nick Oddo"
date: "2024-07-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Problem 5.1
Studying: The files school1.dat, school2.dat and school3.dat contain data on the amount of time students from three high schools spent on studying or homework during an exam period. Analyze data from each of these schools separately, using the normal model with a conjugate prior distribution, in which {$\mu_0 = 5, \sigma_0^2 = 4, \kappa_0 = 1, \nu_0 = 2$} and compute or approximate the following:

a. posterior means and 95% confidence intervals for the mean $\theta$ and standard deviation $\sigma$ from each school;

b. the posterior probability that $\theta_i < \theta_j < \theta_k$ for all six permutations {i,j,k} of {1,2,3};

c. the posterior probability that $\tilde{Y_i} < \tilde{Y_j} <\tilde{Y_k}$ for all six permutations {i, j, k} of {1, 2, 3}, where Y ̃i is a sample from the posterior predictive distribution of school i.

d. Compute the posterior probability that $\theta_1$ is bigger than both $\theta_2$ and $\theta_3$, and the posterior probability that $Y_1$ is bigger than both $Y_2$ and $Y_3$.


```{r, echo = FALSE}
w1 = "http://www2.stat.duke.edu/~pdh10/FCBS/Exercises/school1.dat"
w2 = "http://www2.stat.duke.edu/~pdh10/FCBS/Exercises/school2.dat"
w3 = "http://www2.stat.duke.edu/~pdh10/FCBS/Exercises/school3.dat"

s1 <- read.delim(w1, header = FALSE, sep="\n") 
s2 <- read.delim(w2, header = FALSE, sep="\n") 
s3 <- read.delim(w3, header = FALSE, sep="\n") 

```

Build function to sample from posterior distributions for $\theta$ (mean). `norm.norm.post.theta`

```{r echo = FALSE}

#function to sample from posterior dist. of theta- Peter Hoff parameterization
norm.norm.post.theta <- function(x, k0 ,mu0 ,v0 ,sigma0, num.samples = 1000 ){
  n = length(x)
  xbar = mean(x)
  s2 = var(x)
  

#posterior parameters
#mu|data:
  kn = k0 + n
  mun = (k0*mu0 + n*xbar)/kn

# sigma|data:
    vn = v0 + n
    sigma_n = 1/vn * (v0*sigma0 + (n-1)*s2 + (xbar - mu0)^2*k0*n/kn)

#Sample posterior distributions
  sigma = 1/rgamma(num.samples, vn/2, sigma_n*vn/2)
  mu = rnorm(num.samples, mun, sqrt(sigma_n/kn))

return(mu)
}

```

a. posterior means and 95% confidence intervals for the mean $\theta$ and standard deviation $\sigma$ from each school;

```{r}
num.samples = 10e4

thetaS1 = norm.norm.post.theta(s1$V1, mu0 = 5, sigma0 = 4 , k0 = 1, v0 = 2, num.samples = num.samples)
thetaS2 = norm.norm.post.theta(s2$V1, mu0 = 5, sigma0 = 4 , k0 = 1, v0 = 2, num.samples = num.samples)
thetaS3 = norm.norm.post.theta(s3$V1, mu0 = 5, sigma0 = 4 , k0 = 1, v0 = 2, num.samples = num.samples)

cat(
  paste("school 1 theta mean: "
        , round(mean(thetaS1),3),"; 95% confidence interval: ["
        , round(quantile(thetaS1, .025),3),","
        , round(quantile(thetaS1, .975),3),"]"
        ),
  
  paste("school 2 theta mean: "
        , round(mean(thetaS2),3),"; 95% confidence interval: ["
        , round(quantile(thetaS2, .025),3),","
        , round(quantile(thetaS2, .975),3),"]"
        ),
  
  paste("school 3 theta mean: "
        , round(mean(thetaS3),3),"; 95% confidence interval: ["
        , round(quantile(thetaS3, .025),3),","
        , round(quantile(thetaS3, .975),3),"]"
        )
,sep = '\n')
```

b. the posterior probability that $\theta_i < \theta_j < \theta_k$ for all six permutations {i,j,k} of {1,2,3};

```{r}
library(gtools)
# create data frame that lists permutations.
x = c("thetaS1", "thetaS2", "thetaS3")
p = data.frame(permutations(n = 3, r = 3, v = x), "Prob" = 0.0); colnames(p)<-c("theta_i","theta_j","theta_k","P(theta_i < theta_j < theta_k)")
 
#Calculate probability that theta_i<theta_j<theta_k via monte carlo simulation
for (i in 1:nrow(p) ){
  p[i,4] =   mean( ifelse(get(p[i,1]) < get(p[i,2]) & get(p[i,2])< get(p[i,3]), 1.0,0.0) )
}

p
```

c. the posterior probability that $\tilde{Y_i} < \tilde{Y_j} <\tilde{Y_k}$ for all six permutations {i, j, k} of {1, 2, 3}, $\tilde{Y_i}$ is a sample from the posterior predictive distribution of school i.

```{r echo = FALSE}
#function to sample from posterior predictive dist.
norm.norm.post.pred <- function(x, k0 ,mu0 ,v0 ,sigma0, num.samples = 1000 ){
# posterior predictive distribution:

  # sampling dist ~ N(mu, sigma)
  # posterior dist for sigma ~ 1/gamma(vn, sigma_n)
  # posterior dist for mu ~ N( mun, sigma_n/kn)
  
# N(mu,sigma) + N(mun, sigma_n/kn)

  # y_new ~ N (mu, sigma + sigma_n/kn)

  n = length(x)
  xbar = mean(x)
  s2 = var(x)
  

#posterior parameters
#mu|data:
  kn = k0 + n
  mun = (k0*mu0 + n*xbar)/kn

# sigma|data:
    vn = v0 + n
    sigma_n = 1/vn * (v0*sigma0 + (n-1)*s2 + (xbar - mu0)^2*k0*n/kn)

#Sample posterior distributions
  sigma = 1/rgamma(num.samples, vn/2, sigma_n*vn/2)
  mu = rnorm(num.samples, mun, sqrt(sigma_n/kn))

 post_predictive = rnorm( num.samples, mu, sqrt(sigma + sigma_n/kn) ) 
  
return(post_predictive)
}

```
create funcrion `norm.norm.post.pred` to sample from posterior distribution $\tilde{Y_i}|\mu, \sigma^2$

```{r}
num.samples = 10e4

YtildeS1 = norm.norm.post.pred(s1$V1, mu0 = 5, sigma0 = 4 , k0 = 1, v0 = 2, num.samples = num.samples)
YtildeS2 = norm.norm.post.pred(s2$V1, mu0 = 5, sigma0 = 4 , k0 = 1, v0 = 2, num.samples = num.samples)
YtildeS3 = norm.norm.post.pred(s3$V1, mu0 = 5, sigma0 = 4 , k0 = 1, v0 = 2, num.samples = num.samples)


# create data frame that lists permutations.
x = c("YtildeS1", "YtildeS2", "YtildeS3")
s = data.frame(permutations(n = 3, r = 3, v = x), "Prob" = 0.0); colnames(s)<-c("school_i","school_j","school_k","P(YTilde_i < YTilde_j < YTilde_k)")
 
#Calculate probability that YTilde_i < YTilde_j < YTilde_k via monte carlo simulation
for (i in 1:nrow(s) ){
  s[i,4] =   mean( ifelse(get(s[i,1]) < get(s[i,2]) & get(s[i,2])< get(s[i,3]), 1.0,0.0) )
}

s

```
d. Compute the posterior probability that $\theta_1$ is bigger than both $\theta_2$ and $\theta_3$, and the posterior probability that $Y_1$ is bigger than both $Y_2$ and $Y_3$.

```{r}

m1 = mean( ifelse( thetaS1 > thetaS2 & thetaS1 > thetaS2, 1.0, 0.0 ) )
y1 = mean( ifelse( YtildeS1 > YtildeS2 & YtildeS1 > YtildeS3, 1.0, 0.0 ) )

cat(
  paste( 'P(thetaS1 > thetaS2 and thetaS2) = ',m1 ), 
  paste( 'P(YTildeS1 > YTildeS2 and YTildeS2) = ',y1 ) 
,sep = '\n')

```


# Problem 5.2
Sensitivity analysis: Thirty-two students in a science classroom were
randomly assigned to one of two study methods, A and B, so that
$n_A$ = $n_B$ = 16 students were assigned to each method. After several
weeks of study, students were examined on the course material with an
exam designed to give an average score of 75 with a standard deviation of
10. The scores for the two groups are summarized by ${\bar{y}_A = 75.2, S_A = 7.3}$
and ${\bar{y}_B = 77.5, S_B = 8.1}$. Consider independent, conjugate normal prior
distributions for each of $\theta_A$ and $\theta_B$, with $\mu_0 = 75$ and $\sigma_0^2 = 100$ for
both groups. For each $(\kappa_0, \nu_0) \in {(1,1),(2,2),(4,4),(8,8),(16,16),(32,32)}$
(or more values), obtain $Pr(\theta_A < \theta_B|\mathbf{y}_A, \mathbf{y}_B)$ via Monte Carlo sampling.
Plot this probability as a function of $(\kappa_0 = \nu_0)$. Describe how you might
use this plot to convey the evidence that $\theta_A < \theta_B$ to people of a variety
of prior opinions.

Build function to sample from posterior distributions for $\theta$ (mean). `norm.norm.post.theta`
```{r echo = FALSE}

#function to sample from posterior dist. of theta- Peter Hoff parameterization
norm.norm.post.theta <- function(x, k0 ,mu0 ,v0 ,sigma0,xbar = NULL,n = NULL ,s2 = NULL, num.samples = 1000 ){
  
  if(is.null(xbar)){
  xbar = mean(x)
  }
  
  if(is.null(n)){
  n = length(x)
  }
  
  if(is.null(s2)){
    s2 = var(x)
  }

#posterior parameters
#mu|data:
  kn = k0 + n
  mun = (k0*mu0 + n*xbar)/kn

# sigma|data:
    vn = v0 + n
    sigma_n = 1/vn * (v0*sigma0 + (n-1)*s2 + (xbar - mu0)^2*k0*n/kn)

#Sample posterior distributions
  sigma = 1/rgamma(num.samples, vn/2, sigma_n*vn/2)
  mu = rnorm(num.samples, mun, sqrt(sigma/kn))

  post.samples.list = list( "mu" = mu, "sigma" = sigma)
  
return(post.samples.list)
}

```

sample data using $\theta_0 = 75, \sigma_0^2 = 100$ and a variety of different prior $\sigma^2_0$ specifications.


```{r}

sigma.params = c(1,2,4,8,16,32)

p = data.frame('(k0,v0)'= 0, 'P(thetaA < thetaB)' = 0); colnames(p) = c('(k0,v0)','P(thetaA < thetaB)')


for (i in 1:length(sigma.params)){
  k0 <- v0 <- sigma.params[i]
  
  p[i,1] = paste('(',k0,',',v0,')',sep = '') 

  post.samples.A = norm.norm.post.theta(xbar =75.2, n = 16 ,s2 = (7.3)^2, mu0 = 75 , sigma0 = 100, k0 = k0, v0 = v0)
  post.samples.B = norm.norm.post.theta(xbar =77.5, n = 16 ,s2 = (8.1)^2, mu0 = 75 , sigma0 = 100, k0 = k0, v0 = v0)

  p[i,2] = mean( post.samples.A$mu < post.samples.B$mu ) 
  
}

p

#plot points
plot(
    x=sigma.params,
    y=p$`P(thetaA < thetaB)`,
    type = 'b',
    xlab = 'v0 = k0 = x',
    ylab = "P(thetaA < thetaB)"
    )

```
The prior parameters $(\kappa_0, \nu_0 )$ indicate prior sample size and variance respectively. The higher $\kappa_0$, the more weight will be given to the prior information into the posterior distribution of $\sigma$ (or $\tau=\frac{1}{\sigma}$). The higher $\nu_0$ is the more variation is present in the prior (in other words, the flatter the prior will be). In the case where $\nu_0 = \kappa_0 = i$, the higher $i$, not only will the posterior be more dependent on the prior, but also, the posterior will reflect more variation (also get flatter). The takaway here, however is that regardless of prior selected, it's highly likely that $\theta_A < \theta_B$ since $P(\theta_A < \theta_B) > .5$ for all selected values of $(\kappa_0, \nu_0 )$.

